Generalized spaces for constructive algebra

### 2020-04-06 paper-herrsching2019/main.tex
Noetherianity

Noetherian conditions are notoriously tricky in constructive mathematics; XXX
There are several proposals for constructively sensible definition of
Noetherian rings in the literature on constructive algebra, each with unique
advantages and
disadvantages~\cite{richman:noetherian,mines-richman-ruitenburg:constructive-algebra,perdry:noetherian,perdry:lazy,perdry-schuster:noetherian,tennenbaum:hilbert}.
Insightful comments on why this is so can be found in the introduction and more
specifically on page~27 of the textbook by Henri Lombardi and Claude
Quitté~\cite{lombardi-quitte:constructive-algebra}.

A ring is \emph{anonymously Noetherian} if and only if any of its ideals is
\notnot finitely generated.

%\begin{ex}There is an intuitionistic proof that the ring~$\ZZ$ is anonymously
%Noetherian: Let~$\aaa \subseteq \ZZ$ be any ideal. Under the assumption that
%either there exists a nonzero element in~$\aaa$ or not, the ideal~$\aaa$ is
%\notnot finitely generated, even \notnot principal: For in the first case, a
%minimal element~$d$ of~$\aaa \cap \NN^+$ (which \notnot exists) witnesses~$\aaa
%= (d)$. In the second case the ideal~$\aaa$ is the zero ideal. Since the
%assumption is \notnot satisfied, the ideal~$\aaa$ is \notnot \notnot finitely
%generated, so \notnot finitely generated. (We remark on this proof scheme on
%page~\pageref{proof-scheme-boxed-statements}.) \end{ex}
%
%It appears that this Noetherian condition has not been studied in the
%literature on constructive algebra. Indeed, from the philosophical and the
%computational point of view on constructive mathematics, the notion of an anonymously
%Noetherian ring is not very useful: From an intuitionistic proof that a given
%ideal is finitely generated one can mechanically extract explicit generators,
%thereby satisfying computational or philosophical demands (``the generators are
%really there''). In contrast, an intuitionistic proof that a given ideal is
%\notnot finitely generated doesn't contain computational content in general.
%
%Along the same lines, one could dismiss the fact (proved below) that Hilbert's
%basis theorem, stating that~$A[X]$ is Noetherian if~$A$ is, holds for the
%notion of anonymously Noetherian rings. In fact, one could feel mocked by this
%version of Hilbert's basis theorem: It promises that any ideal of~$A[X]$
%``has'' finitely many generators in some Platonic sense (without providing any
%clue on how one might go on to find the generators -- they remain \emph{anonymous}), provided
%that any ideal of~$A$ ``has'' finitely many generators in the same
%sense.\footnote{We borrowed the term ``anonymous'' from type theory, where it
%is used with a similar meaning (see for
%instance~\cite{kraus-escardo-coquand-altenkirch:anonymous}). However, there is
%a subtle difference: Unique and anonymous-in-the-sense-of-type-theory existence
%implies existence, but \notnot existence does not.}

@@

>>>>

Noetherian conditions are notoriously tricky in constructive algebra, not
least because of their overuse in classical commutative algebra which render
attempts to extract constructive content more challenging.

There are several proposals for constructively sensible definition of
Noetherian rings in the literature on constructive algebra, each with unique
advantages and
disadvantages~\cite{richman:noetherian,mines-richman-ruitenburg:constructive-algebra,perdry:noetherian,perdry:lazy,perdry-schuster:noetherian,tennenbaum:hilbert}.
Insightful comments on why this is so can be found in the introduction and more
specifically on page~27 of the textbook by Henri Lombardi and Claude
Quitté~\cite{lombardi-quitte:constructive-algebra}. To this list of conditions,
we add the following.

\begin{definition}
\label{defn:anonymously-noetherian}
A ring is \emph{anonymously Noetherian} if and only if any of its ideals is
\notnot finitely generated.
\end{definition}

% Ex.: Z finitely generated

This definition is at odds with the idea that constructive mathematics should
be informative, because it only expresses that it is impossible for no finite
generating family to exist without requiring that such a family should be given
-- it may remain \emph{anonymous}.\footnote{We borrowed the term ``anonymous''
from type theory, where it is used with a similar meaning (see for instance
Ref.~\refcite{kraus-escardo-coquand-altenkirch:anonymous}). However, there is a
subtle difference: The conjunction of ``there is at most one element~$x$'' and
``there is anonymously an element~$x$ in the sense of type theory'' implies
that there actually exists an element~$x$. In contrast, \notnot existence does
not.}

However, it is still useful in constructive algebra when interpreted by the
sheaf semantics: If~$A$ is reduced, then~$A^\sim$ is anonymously Noetherian.
XXX

%\begin{ex}There is an intuitionistic proof that the ring~$\ZZ$ is anonymously
%Noetherian: Let~$\aaa \subseteq \ZZ$ be any ideal. Under the assumption that
%either there exists a nonzero element in~$\aaa$ or not, the ideal~$\aaa$ is
%\notnot finitely generated, even \notnot principal: For in the first case, a
%minimal element~$d$ of~$\aaa \cap \NN^+$ (which \notnot exists) witnesses~$\aaa
%= (d)$. In the second case the ideal~$\aaa$ is the zero ideal. Since the
%assumption is \notnot satisfied, the ideal~$\aaa$ is \notnot \notnot finitely
%generated, so \notnot finitely generated. (We remark on this proof scheme on
%page~\pageref{proof-scheme-boxed-statements}.) \end{ex}
%
%It appears that this Noetherian condition has not been studied in the
%literature on constructive algebra. Indeed, from the philosophical and the
%computational point of view on constructive mathematics, the notion of an anonymously
%Noetherian ring is not very useful: From an intuitionistic proof that a given
%ideal is finitely generated one can mechanically extract explicit generators,
%thereby satisfying computational or philosophical demands (``the generators are
%really there''). In contrast, an intuitionistic proof that a given ideal is
%\notnot finitely generated doesn't contain computational content in general.
%
%Along the same lines, one could dismiss the fact (proved below) that Hilbert's
%basis theorem, stating that~$A[X]$ is Noetherian if~$A$ is, holds for the
%notion of anonymously Noetherian rings. In fact, one could feel mocked by this
%version of Hilbert's basis theorem: It promises that any ideal of~$A[X]$
%``has'' finitely many generators in some Platonic sense (without providing any
%clue on how one might go on to find the generators -- they remain \emph{anonymous}), provided
%that any ideal of~$A$ ``has'' finitely many generators in the same
%sense.\footnote{}

@@

\begin{exercise}{Noetherian properties of the integers}%
\begin{enumerate}
\item Show that the law of excluded middle holds if the ring~$\ZZ$ of integers
is Noetherian in the sense that any ideal is finitely generated.\smallskip

{\scriptsize\emph{Hint.} For a truth value~$\varphi$, consider the ideal~$\{ x
\in \ZZ \,|\, x = 0 \vee \varphi \}$.\par}
\item Give an example of a space~$X$ and a subsheaf~$F$ of the constant
sheaf~$\underline{\ZZ}$ such that, under the sheaf semantics, the subsheaf~$F$
is an ideal, and such that it is not the case that, again under the sheaf
semantics, the ideal~$F$ is finitely generated.
\item Show that~$\ZZ$ is anonymously Noetherian in the sense of
Definition~\ref{defn:anonymously-noetherian}.
\end{enumerate}
\end{exercise}
